Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study

### 2020-10-16 manuscript/plos/review_1.tex
addressed comments from fabienne and christine

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Network methods provide a comprehensive approach to uncovering the genetics of complex diseases and building hypotheses. They discover susceptibility genes by jointly consider the statistical association between genetic variation and a phenotype, measured in a genome-wide association study (GWAS), and the biological context of each gene, represented as a network. Hence, a network method could select a gene with a high P-value of association if it connects multiple low P-value genes in the network. In this work, we studied six network methods which identify subnetworks with high scores of GWAS association with a phenotype. This allows them to give more compelling results than standard SNP- and gene-level GWAS analyses, recovering causal subnetworks tightly related to cancer susceptibility. We applied them to GENESIS, a GWAS on French women with a family history of breast cancer and who tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We critically compared these six methods, discussing the impact of their different mathematical frameworks, and the parameter choices. Additionally, we performed an in-depth benchmarking with respect to the size and predictive power of their solutions, as well as their stability and runtimes. Importantly, we found significant overlaps between the genes in five of the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Since most of the methods produced reasonable solutions, we proposed to combine them into a consensus solution, containing the genes selected by at least two methods. This aggregation brought further insights. For instance, it contained \emph{COPS5}, a gene related to multiple hallmarks of cancer, and 14 of its neighbors. The main drawback of network methods with regards to conventional $\chi^2$-based GWAS was instability i.e. network methods' outputs changed more in front of small perturbations in the input. To compensate for this instability, we proposed a stable consensus solution, formed by the most consistently selected genes in different subsamples of the data. This stable consensus was composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than most genes. The network was organized around \emph{CUL3}, which encodes a ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, we showed how network address limitations of GWAS, namely their lack of statistical power and the difficulty of their interpretation. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
In genome-wide association studies (GWAS), thousands of genomes are scanned to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of hereditary cancers. Yet, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. The omnigenic model offers an appealing explanation: only a limited number of core genes are directly involved in the disease; but gene functions are deeply interrelated so that many other genes can alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. In this study, we characterize six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes, and provides a mechanistic explanation for their role in the development of the disease. We provide ready-to-use implementations of all the examined methods.
\linenumbers

@@

In this work, we analyze the application of these six network methods on GWAS data. They use different interpretations of the omnigenic model, and provide a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network methods to recover additional breast cancer susceptibility genes. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus solution of predisposition to familial breast cancer.

@@

\subsection{Conventional SNP- and gene-based analyses confirm that \emph{FGFR2} locus is associated with familial breast cancer}
\label{results:conventional}

@@

Closer examination reveals two other regions (3p24 and 8q24) having low, albeit not genome-wide significant, P\=/values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression using all the GENESIS genotypes as input, and the phenotype (cancer/healthy) as outcome (Section~\ref{methods:classifier}). The algorithm selected 100 SNPs, both from all aforementioned regions and new ones (\nameref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P\=/value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what would be expected under the omnigenic model, the genes to which these SNPs map to (Section~\ref{methods:snp2gene}) are not interconnected in the protein-protein interaction network (PPIN, Section~\ref{methods:networks}). Moreover, the classification performance of the model is low (sensitivity = 55\%, specificity = 55\%, Section \ref{methods:benchmark}). Together, these issues motivate exploring network methods, which consider not only statistical association, but also the location of each gene in a PPIN to find susceptibility genes.

@@

Topologically, the consensus consists of a connected component composed of 49 genes, and multiple smaller subnetworks. Among the latter, 19 genes are in subnetworks containing a single gene or two connected nodes, implying that they do not have a consistently altered neighborhood, but are strongly associated themselves and hence picked by at least two methods. The large connected component contains genes that are highly central in the PPIN, a property which is weakly anti-correlated with the P\=/value of association to the disease (Pearson correlation coefficient~=~-0.26, \nameref{sfig:consensus_stats}). This suggests that these genes were selected because they were on the shortest path between another two highly associated genes. In view of this, we hypothesize that highly central genes might contribute to the heritability through alterations of their neighborhood, consistently with the omnigenic model of disease \cite{boyle_expanded_2017}. For instance, the most central node in the consensus solution is COPS5, a component of the COP9 signalosome which regulates multiple signaling pathways. \emph{COPS5} is related to multiple hallmarks of cancer and is overexpressed in multiple tumors, including breast and ovarian cancer \cite{liu_jab1_cops5_2018}. Despite its lack of association in GENESIS or BCAC (VEGAS2 P\=/value of 0.22 and 0.14 respectively), its neighbors in the consensus solution have consistently low P\=/values (median VEGAS2 P\=/value~=~0.006).

@@

We compared the results obtained with different network methods to the European sample of the Breast Cancer Association Consortium (BCAC) \cite{Michailidou2017}, the largest GWAS to date on breast cancer (Section~\ref{methods:bcac}). Although BCAC case-control studies do not necessarily target cases with a familial history of breast cancer, this comparison is pertinent since we expect a shared genetic architecture at the gene level, at which most network methods operate. This shared genetic architecture, together with BCAC's scale (90 times more samples than GENESIS) provides a reasonable counterfactual of what we would expect if GENESIS had a larger sample size. We computed a gene association score on BCAC (Section~\ref{methods:bcac}). The solutions provided by the different  network methods overlap significantly with BCAC findings (Fisher's exact test P\=/value \textless~0.019). The gene-based methods achieve comparable precision (2\%-25\%) and recall (1.3-12.1\%) at recovering BCAC-significant genes (\nameref{sfig:additional-benchmarks}A). Interestingly, while SConES GI, at the SNP-level, achieves a similar recall (8.6\%), it shows a much higher precision (47.3\%).

@@

Across this article, we have compared our results to significant genes and SNPs in the BCAC study \cite{Michailidou2017}. Network methods show modest precision, but much higher recall at recovering BCAC biomarkers (Section \ref{results:boost}). While precision might be desirable when a subset of good markers is required (for instance, for diagnosis), higher recall is desirable in exploratory settings. Nonetheless, BCAC is not an ideal ground truth.% in which GENESIS had more samples% . Yet, there are caveats which make it a non-optimal comparison. 
First, the studied populations are non-overlapping: BCAC focuses on a pan-European cohort, while GENESIS targets the French population specifically. Second, the study designs differ: a high proportion of breast cancer cases investigated in BCAC are sporadic (not selected according to family history), while GENESIS is a homogeneous dataset not included in BCAC and which focuses on the French high-risk population attending the family cancer clinics. Despite these differences, we expect some degree of shared genetic architecture, especially at the gene level. Finally, and this is indeed the motivation for this study, GWAS are unlikely to identify all genes relevant for the disease: some might only show up in rare-variant studies; others might have too low effect sizes. Network methods account for this by including genes with low association scores but with relevant topological properties. Hence, network methods and GWAS, even well-powered, are unlikely to capture exactly the same sets of genes. This might partly excuse the low precisions displayed in Section \ref{results:boost} and the low AUC displayed in Section \ref{results:benchmark}.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Network methods provide a comprehensive approach to uncovering the genetics of complex diseases and building hypotheses. They discover susceptibility genes by jointly consider the statistical association between genetic variation and a phenotype, measured in a genome-wide association study (GWAS), and the biological context of each gene, represented as a network. Hence, a network method could select a gene with a high P-value of association if it connects multiple low P-value genes in the network. In this work, we evaluate six network methods which identify subnetworks with high scores of GWAS association with a phenotype. This allows them to give more compelling results than standard SNP- and gene-level GWAS analyses, recovering causal subnetworks tightly related to cancer susceptibility. We applied them to GENESIS, a GWAS on French women with a family history of breast cancer and who tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We critically compared these six methods, discussing the impact of their different mathematical frameworks, and the parameter choices. Additionally, we performed an in-depth benchmarking with respect to the size and predictive power of their solutions, as well as their stability and runtimes. Importantly, we found significant overlaps between the genes in five of the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Since most of the methods produced reasonable solutions, we proposed to combine them into a consensus solution, containing the genes selected by at least two methods. This aggregation brought further insights. For instance, it contained \emph{COPS5}, a gene related to multiple hallmarks of cancer, and 14 of its neighbors. The main drawback of network methods with regards to conventional $\chi^2$-based GWAS was instability i.e. network methods' outputs changed more in front of small perturbations in the input. To compensate for this instability, we proposed a stable consensus solution, formed by the most consistently selected genes in different subsamples of the data. This stable consensus was composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than most genes. The network was organized around \emph{CUL3}, which encodes a ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, we showed how network address limitations of GWAS, namely their lack of statistical power and the difficulty of their interpretation. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
In genome-wide association studies (GWAS), thousands of genomes are scanned to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of hereditary cancers. Yet, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. The omnigenic model offers an appealing explanation: only a limited number of core genes are directly involved in the disease; but gene functions are deeply interrelated so that many other genes can alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. In this study, we apply and compare six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes, and provides a mechanistic explanation for their role in the development of the disease. We provide ready-to-use implementations of all the examined methods.
\linenumbers

@@

In this work, we analyze the application of these six network methods on GWAS data. They use different interpretations of the omnigenic model, and provide a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network methods to identify additional breast cancer susceptibility genes. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus solution of predisposition to familial breast cancer.

@@

\subsection{Conventional SNP- and gene-based analyses retrieve the \emph{FGFR2} locus in the GENESIS dataset}
\label{results:conventional}

@@

Closer examination reveals two other regions (3p24 and 8q24) having low, albeit not genome-wide significant, P\=/values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression using all GENESIS genotypes as input, and the phenotype (cancer/healthy) as outcome (Section~\ref{methods:classifier}). The algorithm selected 100 SNPs, both from all aforementioned regions and new ones (\nameref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P\=/value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what would be expected under the omnigenic model, the genes to which these SNPs map to (Section~\ref{methods:snp2gene}) are not interconnected in the protein-protein interaction network (PPIN, Section~\ref{methods:networks}). Moreover, the classification performance of the model is low (sensitivity = 55\%, specificity = 55\%, Section \ref{methods:benchmark}). Together, these issues motivate exploring network methods, which consider not only statistical association, but also the location of each gene in a PPIN to find susceptibility genes.

@@

Topologically, the consensus consists of a connected component composed of 49 genes, and multiple smaller subnetworks. Among the latter, 19 genes are in subnetworks containing a single gene or two connected nodes, implying that they do not have a consistently altered neighborhood, but are strongly associated themselves and hence picked by at least two methods. The large connected component contains genes that are highly central in the PPIN, a property which is weakly anti-correlated with the P\=/value of association to the disease (Pearson correlation coefficient~=~-0.26, \nameref{sfig:consensus_stats}). This suggests that these genes were selected because they were on the shortest path between another two highly associated genes. In view of this, we hypothesize that highly central genes might contribute to the heritability through alterations of their neighborhood, consistently with the omnigenic model of disease \cite{boyle_expanded_2017}. For instance, the most central node in the consensus solution is COPS5, a component of the COP9 signalosome which regulates multiple signaling pathways. \emph{COPS5} is related to multiple hallmarks of cancer and is overexpressed in multiple tumors, including breast and ovarian cancer \cite{liu_jab1_cops5_2018}. Despite its lack of association in GENESIS or in studies conducted by the Breast Cancer Association Consortium (BCAC) \cite{Michailidou2017} (VEGAS2 P\=/value of 0.22 and 0.14 respectively), its neighbors in the consensus solution have consistently low P\=/values (median VEGAS2 P\=/value~=~0.006).

@@

We compared the results obtained with different network methods to the European sample of BCAC, the largest GWAS to date on breast cancer (Section~\ref{methods:bcac}). Although BCAC case-control studies do not necessarily target cases with a familial history of breast cancer, this comparison is pertinent since we expect a shared genetic architecture at the gene level, at which most network methods operate. This shared genetic architecture, together with BCAC's scale (90 times more samples than GENESIS) provides a reasonable counterfactual of what we would expect if GENESIS had a larger sample size. We computed a gene association score on BCAC (Section~\ref{methods:bcac}). The solutions provided by the different  network methods overlap significantly with BCAC findings (Fisher's exact test P\=/value \textless~0.019). The gene-based methods achieve comparable precision (2\%-25\%) and recall (1.3-12.1\%) at recovering BCAC-significant genes (\nameref{sfig:additional-benchmarks}A). Interestingly, while SConES GI, at the SNP-level, achieves a similar recall (8.6\%), it shows a much higher precision (47.3\%).

@@

In this work, we have compared our results to significant genes and SNPs in the BCAC study \cite{Michailidou2017}. Network methods show modest precision, but much higher recall at recovering BCAC hits (Section \ref{results:boost}). While precision might be desirable when a subset of good markers is required (for instance, for diagnosis), higher recall is desirable in exploratory settings. Nonetheless, BCAC is not an ideal ground truth.% in which GENESIS had more samples% . Yet, there are caveats which make it a non-optimal comparison. 
First, the studied populations are non-overlapping: BCAC focuses on a pan-European cohort, while GENESIS targets the French population specifically. Second, the study designs differ: a high proportion of breast cancer cases investigated in BCAC are sporadic (not selected according to family history), while GENESIS is a homogeneous dataset not included in BCAC and which focuses on the French high-risk population attending the family cancer clinics. Despite these differences, we expect some degree of shared genetic architecture, especially at the gene level. Finally, and this is indeed the motivation for this study, GWAS are unlikely to identify all genes relevant for the disease: some might only show up in rare-variant studies; others might have too low effect sizes. Network methods account for this by including genes with low association scores but with relevant topological properties. Hence, network methods and GWAS, even well-powered, are unlikely to capture exactly the same sets of genes. This might partly excuse the low precisions displayed in Section \ref{results:boost} and the low AUC displayed in Section \ref{results:benchmark}.
